Informing policy via dynamic models: Cholera in Haiti

### 2023-08-31 ms-plos.Rnw
Model 3 initialization model fixed
Made most of final changes, need to adjust how supplement and ms interact

# This are initial values for Y_u(0). They were estimated, but because they are
# integers, close to the value of the first observation, it was easiest to fit these
# by hand.
m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # TODO: These were for the original initialization model, may need to fit these again by hand (just for-loop through possible inits.)

@@

mod3_cases <- as.data.frame(t(h3Spat@data))
m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.

@@

The four mechanistic models of Lee et al.~\cite{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
% Here, we have put all three models into a common notational framework.
Translations back to the original notation of Lee et al.~\cite{lee20} are given in Table~S-1 of the supplement. % TODO: Check S-1 table matches.

@@

@@

  & \multirow{2}{*}{${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$}

@@

& &

@@

  & (S15) % TODO: Check

@@

  & (S16) % TODO: Check

@@

  & (S19) % TODO: Check

@@

  & \multirow{2}{*}{(S15)} % TODO: Check

@@

  & \multirow{2}{*}{(S16)} % TODO: Check

@@

  & \multirow{2}{*}{(S19)} % TODO: Check

@@

  & \multirow{2}{*}{$\Iinit^{3,7,10} = (
        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} % TODO: Check the unit numbers
  &  % TODO: \multirow{2}{*}{EQUATION NUMBER}

@@

\caption{\label{tab:params}References to the relevant equation are given in parentheses.

@@

}

@@

While the complete model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplement.

@@

Per-capita transition rates are used for both compartments representing human counts and the aquatic reservoir of bacteria; these rates  are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:IR}
\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
\\%[-3pt]
\label{eq:model3:IS}
\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
\\%[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\%[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\%[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
\\%[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in Eqs.~\myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In Eq.~\myeqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in Eq.~\myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in Eqs.~\myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

@@

\begin{table}[!h]
\centering

\begin{tabular}{|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  $\Sexpr{myround(mod3_ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ &
  $\Sexpr{as.character(mod2_n_params)}$ &
  $\Sexpr{as.character(mod3_n_params)}$ \\
 Fit Parameters &
 (\Sexpr{as.character(mod1_lee_n_params)}) &
 ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
 ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ &
  $\Sexpr{myround(mod2_aic, 1)}$ &
  $\Sexpr{myround(mod3_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
\\
\hline
\thead{Benchmark \\ AIC} &
  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}AIC values for each model compared to their respective benchmarks.
Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.  $\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model. $\textsuperscript{2}$ Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
}
\end{table}

@@

% TODO: Remove the unused code below.
% % Place tables after the first paragraph in which they are cited.
% \begin{table}[!ht]
% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
% \centering
% \caption{
% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
% \begin{tabular}{|l+l|l|l|l|l|l|l|}
% \hline
% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
% \end{tabular}
% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
% \end{flushleft}
% \label{table1}
% \end{adjustwidth}
% \end{table}

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

@@

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

>>>>

# This are initial values for Y_u(0). They were estimated, but because they are
# integers, close to the value of the first observation, it was easiest to fit these
# by hand.
m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # fit by hand

@@

mod3_cases <- as.data.frame(t(h3Spat@data))
m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # fit by hand

@@

The four mechanistic models of Lee et al.~\cite{lee20} were deliberately developed with limited coordination.
This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
% Here, we have put all three models into a common notational framework.
Translations back to the original notation of Lee et al.~\cite{lee20} are given in the supplement (\nameref{S1_Table}).

@@

\caption{\label{tab:params} {\bf Model parameters.}}

@@

  & ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$

@@

& ${\fixed \choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}$

@@

  & (S15)

@@

  & (S16)

@@

  & (S19)

@@

  & \multirow{2}{*}{(S15)}

@@

  & \multirow{2}{*}{(S16)}

@@

  & \multirow{2}{*}{(S19)}

@@

  & \multirow{2}{*}{$\Iinit^{3,4} = (
        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H4', 'Iinit4')]))})^*$}
  &  \multirow{2}{*}{(S20)}

@@

\begin{flushleft} References to the relevant equation are given in parentheses.

@@

@@

While the complete model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.
Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplemental text (\nameref{S1_Text}).

@@

Per-capita transition rates are used for both compartments representing human counts and the aquatic reservoir of bacteria; these rates  are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:IR}
\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
\\%[-3pt]
\label{eq:model3:IS}
\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
\\%[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\%[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\%[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t)\big)^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
\\%[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in Eqs.~\myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In Eq.~\myeqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in Eq.~\myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in Eqs.~\myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

@@

\begin{table}[!h]
\centering
\caption{\label{tab:likes}{\bf AIC values for Models~1--3 and their benchmarks.}}
\begin{tabular}{|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  $\Sexpr{myround(mod3_ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ &
  $\Sexpr{as.character(mod2_n_params)}$ &
  $\Sexpr{as.character(mod3_n_params)}$ \\
 Fit Parameters &
 (\Sexpr{as.character(mod1_lee_n_params)}) &
 ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
 ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ &
  $\Sexpr{myround(mod2_aic, 1)}$ &
  $\Sexpr{myround(mod3_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
\\
\hline
\thead{Benchmark \\ AIC} &
  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
\\
\hline
\end{tabular}
\begin{flushleft}
Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.  \textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model. \textsuperscript{2}Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
\end{flushleft}
\end{table}

@@

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Table}
\label{S1_Table}
{\bf Notation conversion table.} Conversions between the notation used here and the notation of Lee et al. (2020) \cite{lee20}.

\paragraph*{S1 Text}
\label{S1_Text}
{\bf Supplemental Material.} In depth description of the numeric implementation of models.

@@

% \paragraph*{S1 Table.}
% \label{S1_Table}
% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
